Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-09 paper/paper.tex
Restored cut out contents

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
[MOVED to background]
\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
(Section~\ref{sec:knap}), and examples of the shortcut deforestation
optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

>>>>

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
\dcas{Do not cut below}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{A} (the
\emph{carrier} of the algebra) together with some operations on it sometimes
subject to some equational laws. For example, given such a type \coq{A} we can
define the monoid operations as \coq{u : unit -> A} for unit
and \coq{m : A * A -> A} for multiplication. Notice
that \coq{F X = unit + X * X} is a functor on \coq{X} which means we can
equivalently describe these two maps as a single one of type \coq{F X -> X}
which we call the algebra for the functor \coq{F} satisfying the unit and
associative laws of the monoid. In this work we will be using algebras to
describe the operations of a data type and so we will not be needing additional
equations on these operations.

Given a type \coq{A} and a functor \coq{F}, an \coq{F}-algebra is a pair given
by a type \coq{A} called the \emph{carrier} of a \emph{structure map} of
type \coq{App F A ~> A}:
\begin{coqcode}
Notation Alg F A := (App F A ~> A).
\end{coqcode}
The least fixed-point for a functor \coq{F} is an instance of an \coq{F}-algebra
where the structure map is an isomorphism. This is sometimes referred to as the
\emph{initial algebra} for a functor \coq{F}. We will explain the reason behind this name shortly. \footnote[1]{To be more precise,
initial algebras are isomorphisms by Lambek's lemma, but isomorphisms of
functors do not necessarily correspond to initial algebras.}  We define the
least fixed-point of \coq{F} as an inductive type:
\begin{coqcode}
Inductive LFix `(F : Cont Sh P) : Type := LFix_in { LFix_out : App F (LFix F) }.
\end{coqcode}
where \coq{LFix_in} is the \coq{F}-algebra while \coq{LFix_out} is its inverse.
As an example, the initial \coq{F}-algebra for the
functor \coq{F X = unit + A * X} is the type of lists with the \coq{F}-algebra
being defined by the empty list \coq{Empty : unit -> LFix F} and the cons
operation \coq{Cons : A * LFix F -> LFix F}.

We define \coq{LFix} as a setoid, where its equivalence relation can be
described as the least fixed point of \coq{AppR} and we define smart
constructors for the isomorphism of least fixed points as respectful morphisms:
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F                l_out : LFix F ~> App F (LFix F)
\end{coqcode}

The least fixed-point for \coq{F} is the initial \coq{F}-algebra in the sense
that it gives rise to an inductive recursion scheme.  Specifically, for any
other \coq{F}-algebra there exists a \emph{unique} map, known as a \emph{fold} or
\emph{catamorphism}, such that it structurally deconstructs the data type
using \coq{LFix_out}, calls itself recursively and then composes the result of
the recursive call using the given an \coq{F}-algebra. In other words, for any
give \coq{F}-algebra there exists a unique \coq{F}-algebra homomorphism from the
initial one. We define it in Coq as follows:
\begin{coqcode}
  Definition cata_f `{F : Cont Sh P} (alg : Alg F A) : LFix -> A
:= fix f (x : LFix) := match x with
  | LFix_in ax => let (sx, kx) := ax in alg (MkApp sx (fun e => f (kx e))) end.
\end{coqcode}
It is easy to show that this function is a respectful morphism of
\coq{F}-algebras. In fact, it is possible to define it as a map of the
following type:
\begin{coqcode}
  cata : forall `{F : Cont Sh P} `{setoid A}, Alg F A ~> LFix ~> A
\end{coqcode}
We prove that catamorphisms satisfy the universality property we explained previously:
\begin{coqcode}
Lemma cata_univ `{eA : setoid A} (alg : Alg F A) (f : LFix ~> A)
    : f =e cata alg <-> f =e alg \o fmap f \o l_out.
\end{coqcode}
In other words, if there is any other \coq{f} with the same structural recursive
shape as the catamorphism on the algebra \coq{alg} then it must be equal to that
catamorphism.

\subsection{Coalgebras and Anamorphisms}
\label{sec:coalg}
Algebras and catamorphisms dualise nicely to the coinductive setting.  The dual
of an algebra is a coalgebra. A coalgebra can be thought of as an
\emph{observation} map. For example, for a type of states \coq{X} and an
alphabet type \coq{L} we can define a labelled transition system (LTS)
on \coq{X} as a function \coq{c : X -> FX} for a functor \coq{F X = L * X}
implementing the \emph{transition} map. In particular, for a
state \coq{x1 : X}, \coq{c x1} returns a pair \coq{(l, x2) : L * X}
where \coq{l : L} is the observable action and \coq{x2 : X} is the next state.

In general, for a functor \coq{F}, an \coq{F}-coalgebra is a pair of a
type \coq{X} called the \emph{carrier} of the coalgebra and a structure
map \coq{X -> F X}.

In our development we use the following notation for coalgebras:
\begin{coqcode}
Notation Coalg F A := (A ~> App F A).
\end{coqcode}

Dually to the initial \coq{F}-algebra, a final \coq{F}-coalgebra is the greatest
fixed-point for a functor \coq{F}. We define it using a coinductive data type:
\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}
where \coq{GFix_out} is the final \coq{F}-coalgebra and \coq{GFix_in} is its
inverse witnessing the isomorphism. Similarly to \coq{LFix}, \coq{GFix} is also
defined as a setoid, with an equivalence relation that is the greatest fixpoint
of \coq{AppR}. Additionally, we define smart constructors for the isomorphism of
greatest fixed points
\begin{coqcode}
g_in : App F (GFix F) ~> GFix F                g_out : GFix F ~> App F (GFix F)
\end{coqcode}

The greatest fixed-point is a terminal \coq{F}-coalgebra in the sense that it
yields a coinductive recursion scheme. Specifically, for any
other \coq{F}-coalgebra there exists a \emph{unique} map, called the
\emph{unfold} or \emph{anamorphism}, such that it applies the observation map,
corecursively generates the rest of the computation and composes the result of
the corecursive call by using the algebra \coq{GFix_in}. In other words, for any
given \coq{F}-coalgebra there exists a unique \coq{F}-coalgebra homomorphism
into the terminal \coq{F}-coalgebra.

% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
% \[\begin{tikzcd}
%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
% 	\arrow["a", from=1-1, to=1-2]
% 	\arrow["b", from=1-2, to=1-3]
% \end{tikzcd}\]
% Note that for a generic functor \coq{F} an \coq{F}-coalgebra on the
% carrier \coq{X} is a map \coq{X -> F X}. In the previous example \coq{c} is
% an \coq{F}-coalgebra on the carrier \coq{X} with \coq{F X = L * X}.

We define anamorphisms as follows:
\begin{coqcode}
Definition ana_f_ (c : Coalg F A) :=
  cofix f x :=
    match c x with | MkApp sx kx => GFix_in (MkApp sx (fun e => f (kx e))) end.

Definition ana : forall `{setoid A}, Coalg F A ~> A ~> GFix F := (*... ana_f_ ... *)
\end{coqcode}
From this definition the universality property falls out:
\begin{coqcode}
Lemma ana_univ `{eA : setoid A} (h : Coalg F A) (f : A ~> GFix F)
: f =e ana h <-> f =e g_in \o fmap f \o h.
\end{coqcode}
In words, for any \coq{F}-coalgebra, if there is any other function \coq{f}
that is a \coq{F}-coalgebra homomorphism then it must be the anamorphism on the
same coalgebra.

\subsection{Recursive Coalgebras and Hylomorphism}
\label{sec:rec-coalgebras}
Hylomorphisms capture the concept of \emph{divide-and-conquer} algorithms where
the input is first destructured (\emph{divide}) in smaller parts by means of a
coalgebra which are computed recursively and then composed back together
(\emph{conquer}) by means of an algebra.

\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
(Section~\ref{sec:knap}), and examples of the shortcut deforestation
optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}
